WORCS: A Workflow for Open Reproducible Code in Science

### 2020-11-14 paper/Manuscript_DS.Rmd
Better address FAIR principles

  affiliation: | 
      a: Utrecht University faculty of Social and Behavioral Sciences, department of Methodology & Statistics  
      b: Open Science Community Utrecht  
      c: Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany  
      d: Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany and London, UK  
      e: UMC Utrecht  
      f: Utrecht University, Department of Information and Computing Sciences  
      g: Utrecht University faculty of Humanities, department of Language, Literature & Communication, Utrecht Institute of Linguistics OTS  
      h: University Library Utrecht
title: "WORCS: A Workflow for Open Reproducible Code in Science"
short:      "WORKFLOW FOR OPEN SCIENCE"
shorttitle:      "WORKFLOW FOR OPEN SCIENCE"
abstract: |
  Adopting open science
  principles can be challenging and time-intensive, because doing so requires
  substantial conceptual education and training in the use of new tools.
  This paper introduces the Workflow for
  Open Reproducible Code in Science (WORCS): A step-by-step
  procedure that researchers can follow to make a research project open and
  reproducible.
  The purpose of the workflow is to lower the threshold for 
  adoption of open science principles.
  It is based on best practices, and can be used either in
  parallel to, or in absence of, top-down requirements by journals,
  institutions, and funding bodies.
  To facilitate widespread adoption, the WORCS principles have been implemented
  in the R package `worcs`, which offers an RStudio 
  project template and utility functions for specific workflow steps.
  This paper introduces the conceptual workflow,
  discusses how it meets different 
  standards for open science, and provides examples of the implementation in `R`. 
  This paper is primarily targeted towards
  scholars conducting research projects in R on a private computer,
  conducting research that involves academic prose, analysis code,
  and (optionally) tabular data.
  However, the workflow is flexible enough to accommodate other scenarios,
  and offers a sensible starting point for customized solutions.
  The source code for the `R` package and manuscript,
  and a list of user examples of WORCS projects, are available at 
  https://github.com/cjvanlissa/worcs.
authornote: "blabla"
keywords:
  # at least one keyword must be supplied
  formatted: [open science, reproducibility, r, dynamic document generation, version control]
  plain:     [open science, reproducibility, r, dynamic document generation, version control]
preamble: >
  \usepackage{amsmath}
bibliography      : ["worcs.bib"]  
urlcolor          : blue
output: rmarkdown::html_document
---
<!--
, "worcs_packages.bib"
output: 
  rmarkdown::html_document:
    citation_package: biblatex Note: Citations are always handled by pandoc when format is not PDF
knit: worcs::cite_all
csl: apa.csl
csl: apa.csl
-->

@@

@@

Many researchers are motivated to adopt current best practices for open
science and enjoy these benefits. And yet, the question of how to 
get started can be daunting. Making the transition
requires researchers to become knowledgeable about different open science 
challenges and solutions, and to become proficient with new and unfamiliar tools.
This paper is designed to ease that transition by presenting a simple workflow,
based on best practices,  
that meets most requirements open science:
The *Workflow for Open Reproducible Code in Science* (WORCS).

This paper introduces the conceptual workflow (referred to in upper case, "WORCS"),
and discusses its underlying principles and how it meets different standards for open science.
The principles underlying this workflow are universal and largely platform independent.
We also present a software implementation of WORCS for `R` users [@rcoreteamLanguageEnvironmentStatistical2020]:
The `R` package `worcs` [referred to in monospace font, @worcspackage2020].
This package offers a project template for RStudio
[@rstudioteamRStudioIntegratedDevelopment2015],
and several convenience functions to automate specific workflow steps.
This paper is most relevant for scholars using R on a private computer to
conducting research that involves academic prose, analysis code,
and tabular data.
For other readers, it can serve as a primer on open science workflows,
and provide a sensible starting point for a customized solution.

The WORCS workflow (Figure 1) 
constitutes a step-by-step procedure that researchers can follow to make a 
research project open and reproducible.
WORCS is compatible with open science
requirements already implemented by journals and institutions, and will help 
fulfill them.
It can also be used by individual authors to produce work in accordance with best practices
in the absence of top-down support or guidelines (i.e., "grass roots" open science).

@@

An integral part of WORCS is the *explicit* version control software Git
(www.git-scm.com).
Git tracks changes to files on a line-by-line basis.
The user can sture these changes by making a "commit": a 
snapshot of the version controlled files.
Each "commit" can contain as many changes as desired;
for example, a whole new paragraph, or many small changes made to address a single reviewer comment.
Each commit is further tagged with a message, describing the changes.
The `R` implementation of `worcs` contains a convenience function,
`git_update("your commit message")`,
which creates a new commit for all changes since the previous commit,
labels it with "your commit message", and pushes these changes to the remote repository.
It is, effectively, a Git "quick save" command.

From an open science perspective, Git is particularly appealing because it
retains a complete historical backlog of all commits.
This can be used, for example, to verify that authors executed the steps outlined in a preregistration.
Users can also compare changes between different commits,
or go back to a previous version of the code (for example,
after making a mistake, or to replicate a previous version of the results).
A project version controlled with Git is called a "Repository", or Repo.
Git only version controls files explicitly committed by the user. Moreover, it
is possible to prevent files from being version controlled - which is useful for
privacy sensitive data.
A `.gitignore` file lists files that should not be version-controlled.
These files thus exist only on the user's private computer.

@@

WORCS was designed to meet the TOP-guidelines, and it does so in a way that
is compatible with other open science principles. Most notably, WORCS is 
amenable to the
FAIR Guiding Principles for scientific data management and stewardship [@wilkinsonFAIRGuidingPrinciples2016a]. These principles advocate that digital research objects should be Findable, Accessible, Interoperable and Reusable. Initially mainly promoted as principles for data, they are increasingly applied as a standard for other types of research output as well,
most notably software [@lamprechtFAIRPrinciplesResearch2019].

WORCS helps users meet the FAIR principles by making all created research objects -
including data and code - accessible in a repository that can be made public. 
The `worcs` package additionally
facilitates meeting the FAIR principles in several ways. First, each `worcs`
project has a YAML file (https://yaml.org/) that lists the research objects in
the project, which will allow web crawlers 
to index `worcs` projects. Second, a codebook is automatically generated for
each data file stored
using `open_data()` or `closed_data()`. We encourage users to elaborate on the
default codebook by adding
variable description and categories, which would - in the future - enable
indexing data by topic area.

@@

Sharing all research objects, as advocated in WORCS, also provides a proper
basis for research evaluation
according to the San Francisco Declaration on Research Assessment (DORA;
https://sfdora.org/), which plays an increasing role in grant funding, hiring, 
and promotion procedures. Direct access to research objects allows stakeholders
to evaluate research quality based on content rather than relying on spurious
surrogate indicators like journal impact factors, conference rankings, and
h-indexes. The detailed version control and commit tracking of Git remote repositories
furthermore make it possible to assess the relative
contributions made by different researchers.

@@

WORCS provides a user-friendly and
lightweight workflow for open, reproducible research, that meets all
TOP-guidelines, except replication.
Nevertheless, there are clear directions for future developments.
Firstly, although the workflow is currently implemented only in `R`, it is 
conceptually relevant for users of other statistical programming languages, and we welcome efforts to implement WORCS in other platforms.
We welcome efforts to implement WORCS in other platforms.
Secondly, even when a project is in `R`, it may have dependencies outside of the
`R` environment that cannot be managed using `renv`.
It is beyond the scope of the `worcs` package to support tools outside of `R`,
or to containerize a project so that it can
be identically reinstated on a different system or virtual machine.
The forthcoming package `repro` [@repro] will offer such extensions of the
workflow, thus combining the strengths of WORCS with the solutions proposed
by @peikertReproducibleDataAnalysis2019.

>>>>

  affiliation: | 
      a: Utrecht University faculty of Social and Behavioral Sciences, department of Methodology & Statistics  
      b: Open Science Community Utrecht  
      c: Center for Lifespan Psychology, Max Planck Institute for Human Development, Berlin, Germany  
      d: Max Planck UCL Centre for Computational Psychiatry and Ageing Research, Berlin, Germany and London, UK  
      e: UMC Utrecht  
      f: Utrecht University, Department of Information and Computing Sciences  
      g: Utrecht University faculty of Humanities, department of Language, Literature & Communication, Utrecht Institute of Linguistics OTS  
      h: University Library Utrecht
title: "WORCS: A Workflow for Open Reproducible Code in Science"
short:      "WORKFLOW FOR OPEN SCIENCE"
shorttitle:      "WORKFLOW FOR OPEN SCIENCE"
abstract: |
  Adopting open science
  principles can be challenging and time-intensive, because doing so requires
  substantial conceptual education and training in the use of new tools.
  This paper introduces the Workflow for
  Open Reproducible Code in Science (WORCS): A step-by-step
  procedure that researchers can follow to make a research project open and
  reproducible.
  The purpose of the workflow is to lower the threshold for 
  adoption of open science principles.
  It is based on established best practices, and can be used either in
  parallel to, or in absence of, top-down requirements by journals,
  institutions, and funding bodies.
  To facilitate widespread adoption, the WORCS principles have been implemented
  in the R package `worcs`, which offers an RStudio 
  project template and utility functions for specific workflow steps.
  This paper introduces the conceptual workflow,
  discusses how it meets different 
  standards for open science, and addresses the functionality provided by the
  `R` implementation, `worcs`. 
  This paper is primarily targeted towards
  scholars conducting research projects in R,
  conducting research that involves academic prose, analysis code,
  and (optionally) tabular data.
  However, the workflow is flexible enough to accommodate other scenarios,
  and offers a sensible starting point for customized solutions.
  The source code for the `R` package and manuscript,
  and a list of [user examples of WORCS projects](https://github.com/cjvanlissa/worcs/tree/master#adoption-of-worcs-by-users), are available at 
  https://github.com/cjvanlissa/worcs.
authornote: "blabla"
keywords:
  # at least one keyword must be supplied
  formatted: [open science, reproducibility, r, dynamic document generation, version control]
  plain:     [open science, reproducibility, r, dynamic document generation, version control]
preamble: >
  \usepackage{amsmath}
bibliography      : ["worcs.bib"]  
urlcolor          : blue
output: rmarkdown::html_document
---
<!--
, "worcs_packages.bib"
output: 
  rmarkdown::html_document:
    citation_package: biblatex Note: Citations are always handled by pandoc when format is not PDF
knit: worcs::cite_all
csl: apa.csl
csl: apa.csl
-->

@@

# Introduction

@@

Many researchers are motivated to adopt current best practices for open
science and enjoy these benefits. And yet, the question of how to 
get started can be daunting. Making the transition
requires researchers to become knowledgeable about different open science 
challenges and solutions, and to become proficient with new and unfamiliar tools.
This paper is designed to ease that transition by presenting a simple workflow,
based on established best practices,  
that meets most requirements open science:
The *Workflow for Open Reproducible Code in Science* (WORCS).

This paper introduces the conceptual workflow (referred to in upper case, "WORCS"),
and discusses its underlying principles and how it meets different standards for open science.
The principles underlying this workflow are universal and largely platform independent.
We also present a software implementation of WORCS for `R` users [@rcoreteamLanguageEnvironmentStatistical2020]:
The `R` package `worcs` [referred to in monospace font, @worcspackage2020].
This package offers a project template for RStudio
[@rstudioteamRStudioIntegratedDevelopment2015],
and several convenience functions to automate specific workflow steps.
This paper is most relevant for scholars using R to
conduct research that involves academic prose, analysis code,
and tabular data.
For other readers, it can serve as a primer on open science workflows,
and provide a sensible starting point for a customized solution.

The WORCS workflow constitutes a step-by-step procedure that researchers can follow to make a 
research project open and reproducible.
See Figure 1 for a schematic overview of the conceptual workflow,
which shows the steps to take in different stages of the research process.
The workflow is elaborated in greater detail below.
WORCS is compatible with open science
requirements already implemented by journals and institutions, and will help 
fulfill them.
It can also be used by individual authors to produce work in accordance with best practices
in the absence of top-down support or guidelines (i.e., "grass roots" open science).

@@

An integral part of WORCS is the *explicit* version control software Git
(www.git-scm.com).
Git tracks changes to files on a line-by-line basis.
The user can store these changes by making a "commit": a 
snapshot of the version controlled files.
Each "commit" can contain as many changes as desired;
for example, a whole new paragraph, or many small changes made to address a single reviewer comment.
Each commit is further tagged with a message, describing the changes.
The `R` implementation of `worcs` contains a convenience function,
`git_update("your commit message")`,
which creates a new commit for all changes since the previous commit,
labels it with "your commit message", and pushes these changes to the remote repository.
It is, effectively, a Git "quick save" command.

From an open science perspective, Git is particularly appealing because it
retains a complete historical backlog of all commits.
This can be used, for example, to verify that authors executed the steps outlined in a preregistration.
Users can also compare changes between different commits,
or go back to a previous version of the code (for example,
after making a mistake, or to replicate a previous version of the results).
A project version controlled with Git is called a "Repository", or Repo.
Git only version controls files explicitly committed by the user. Moreover, it
is possible to prevent files from being version controlled - which is useful for
privacy sensitive data.
A `.gitignore` file lists files that should not be version-controlled.
These files thus exist only on the user's computer.

@@

Aside from the TOP Guidelines, the FAIR Guiding Principles for scientific data management and stewardship [@wilkinsonFAIRGuidingPrinciples2016a] are an important standard for open science.
These principles advocate that digital research objects should be Findable, Accessible, Interoperable and Reusable.
Initially mainly promoted as principles for data, they are increasingly applied as a standard for other types of research output as well,
most notably software [@lamprechtFAIRPrinciplesResearch2019].

WORCS was not designed to comprehensively address these principles, but the workflow does facilitate meeting them.
In terms of Findability, all files in GitHub projects are searchable by content and meta-data, both on the platform and through standard search engines.
The workflow further recommends that users create a project page on the Open Science Framework – a recognized repository in the social sciences – and generate a DOI for this project.
Additional DOIs can be generated through the OSF for specific resources, such as data sets.
It is additionally or alternatively possible to connect the GitHub repository to Zenodo, which also offers DOIs and allows users to document projects with extensive metadata.
Finally, each worcs project has a [YAML file] (https://yaml.org/) that lists the data objects (and other meta-data) in the project, which will allow web crawlers to index worcs projects.
We use this metadata, for example, to identify public `worcs` projects on GitHub.
Accessibility is primarily safeguarded by hosting all project files in a public GitHub repository, which can be downloaded as a ZIP archive, and cloned or forked using the Git protocol. [GitHub is committed to long term accessibility of these resources]( https://archiveprogram.github.com/). We further encourage users to make their data accessible when possible.
When data must be closed, the `worcs` function `closed_data()` adds a paragraph to the project readme, indicating how readers can contact the author for access.

With regard to interoperability, the workflow recommends using plain-text files whenever possible, which ensures that code and meta-data are human- and machine-readable.
Furthermore, the `worcs` package is implemented in R, which is open source.
To promote reusability, we recommend users to specify an appropriate license for their projects.
The `worcs` package includes the CC licenses, and suggests a CC-BY 4.0 license by default.
This license is featured in the GitHub repository, and we encourage users to also display it on their project’s OSF page.
To further facilitate reusability, `worcs` automatically generates a codebook for
each data file.
We encourage users to elaborate on the default codebook by adding
variable description and categories, which would - in the future - enable
indexing data by topic area.

@@

Sharing all research objects, as advocated in WORCS, also provides a thorough
basis for research evaluation
according to the San Francisco Declaration on Research Assessment (DORA;
https://sfdora.org/), which plays an increasing role in grant funding, hiring, 
and promotion procedures. Direct access to research objects allows stakeholders
to evaluate research quality based on content rather than relying on spurious
surrogate indicators like journal impact factors, conference rankings, and
h-indexes. The detailed version control and commit tracking of Git remote repositories
furthermore make it possible to assess the relative
contributions made by different researchers.

@@

WORCS provides a user-friendly and
lightweight workflow for open, reproducible research, that meets all
TOP-guidelines pertaining to reproducibility (1-7).
Nevertheless, there are clear directions for future developments.
Firstly, although the workflow is currently implemented only in `R`, it is 
conceptually relevant for users of other statistical programming languages, and we welcome efforts to implement WORCS in other platforms.
We welcome efforts to implement WORCS in other platforms.
Secondly, even when a project is in `R`, it may have dependencies outside of the
`R` environment that cannot be managed using `renv`.
It is beyond the scope of the `worcs` package to support tools outside of `R`,
or to containerize a project so that it can
be identically reinstated on a different system or virtual machine.
The forthcoming package `repro` [@repro] will offer such extensions of the
workflow, thus combining the strengths of WORCS with the solutions proposed
by @peikertReproducibleDataAnalysis2019.
